Finch: Sparse and Structured Array Programming with Control Flow

### 2024-03-25 sections/tensors.tex
cool

    In Finch, we have identified 8 key level structures that can represent all
    of the relevant combinations of these properties, summarized in Table
    \ref{tab:TypesOfStructure}. We examine each structure in turn, describing
    some of the key properties and potential use cases of each. In this sense,
    the structures we consider are exhaustive. We can represent a wide variety
    of hierarchical tensor structures by combining these level structures in a
    tree, as shown in Figure~\ref{fig:levels}.
    \begin{table}
        \centering
        \begin{tabular}{|c|c|c|c|l|}
            \hline
            \textbf{Sparse} & \textbf{Blocked} & \textbf{Runs} & \textbf{Singletons} & \textbf{Corresponding Format} \\
            \hline
             &  &  &  & Dense \\
            \hline
             &  & \checkmark &  & DenseRLE \\
            \hline
            \checkmark &  &  &  & Sparse \\
            \hline
            \checkmark &  &  & \checkmark & SparsePinpoint \\
            \hline
            \checkmark &  & \checkmark &  & SparseRLE \\
            \hline
            \checkmark &  & \checkmark & \checkmark & SparseInterval \\
            \hline
            \checkmark & \checkmark &  &  & SparseVBL \\
            \hline
            \checkmark & \checkmark &  & \checkmark & SparseBand \\
            \hline
        \end{tabular}
        \caption{All combinations of relevant structural properties and their
        corresponding formats.  Note that blocks and runs need not be considered
        together because we must store a run length for each run, and so there
        isn't a significant storage benefit to combining them. Blocks and
        singletons only make sense in the contex of sparsity, so we don't
        consider them together either. We omit such combinations
        from the exhaustive table.}
        \label{tab:TypesOfStructure}
    \end{table}

    \begin{figure}
        \centering
        %\includegraphics[width=0.45\linewidth]{structuraldiversity.png}\hfill%
        \caption{This figure is a placeholder, but will eventually contain some very beguiling trees of levels.}
        \label{fig:structuraldiversity}
    \end{figure}

@@

Our simplified view of a level is enabled by our use of looplets to represent
the structure within each fiber. In fact, our level interface requires only
5 highly general operations, described below:

Our view of a level as a fiber allocator implies an allocation function
$assemble(tns, pos_{start}:pos_{stop})$, which allocates fibers at positions
$pos_{start}:pos_{stop}$ in the level. We don't specify a de-allocation function,
instead relying on initialization to reset the fiber if it needs to be reused.

\paragraph{$declare(lvl, init, dims...)$} Declares the level to hold subtensors of size $dims$ and an initial value of $init$. 
\paragraph{$declare(lvl, init, dims...)$} Declares the level to hold subtensors of size $dims$ and an initial value of $init$. 
\paragraph{$freeze(lvl, init, dims...)$} Declares the level to hold subtensors of size $dims$ and an initial value of $init$.

\subsection{Core Level Language Primitives}
\begin{enumerate}
\item SparseList
\item SparseDict
\item ...
\end{enumerate}

>>>>

    In Finch, we have identified 8 key level structures that can represent all
    of the relevant combinations of these properties, summarized in Table
    \ref{tab:TypesOfStructure}. We examine each structure in turn, describing
    some of the key properties and potential use cases of each. In this sense,
    the structures we consider are exhaustive. We can represent a wide variety
    of hierarchical tensor structures by combining these level structures in a
    tree, as shown in Figure~\ref{fig:structuraldiversity}.
    \begin{table}
        \centering
        \begin{tabular}{|c|c|c|c|l|}
            \hline
            \textbf{Sparse} & \textbf{Blocked} & \textbf{Runs} & \textbf{Singletons} & \textbf{Corresponding Format} \\
            \hline
             &  &  &  & Dense \\
            \hline
             &  & \checkmark &  & DenseRLE \\
            \hline
            \checkmark &  &  &  & Sparse \\
            \hline
            \checkmark &  &  & \checkmark & SparsePinpoint \\
            \hline
            \checkmark &  & \checkmark &  & SparseRLE \\
            \hline
            \checkmark &  & \checkmark & \checkmark & SparseInterval \\
            \hline
            \checkmark & \checkmark &  &  & SparseVBL \\
            \hline
            \checkmark & \checkmark &  & \checkmark & SparseBand \\
            \hline
        \end{tabular}
        \caption{All combinations of relevant structural properties and their
        corresponding formats.  Note that blocks and runs need not be considered
        together because we must store a run length for each run, and so there
        isn't a significant storage benefit to combining them. Blocks and
        singletons only make sense in the contex of sparsity, so we don't
        consider them together either. We omit such combinations
        from the exhaustive table.}
        \label{tab:TypesOfStructure}
    \end{table}

    \begin{figure}
        \centering
        \includegraphics[width=\linewidth]{Structures.png}\hfill%
        \caption{Several examples of matrix structures represented using the
        level structures identified in Table~\ref{tab:TypesOfStructure}.
        Comparing this figure to \cite[Figure 3]{ahrens_looplets_2023}, we see
        that a level-by-level structural decomposition is diagrammed together
        with the Looplets.}
        \label{fig:structuraldiversity}
    \end{figure}

@@

Our simplified view of a level is enabled by our use of looplets to represent
the structure within each fiber. In fact, our level interface requires only 5
highly general operations, described below.

The first three of these functions, declare, freeze, and thaw, have to do with
managing when tensors can be assumed mutable or immutable. As we use Looplets to
represent iteration over a tensor, we must restrict the mutability of tensors
while we iterate over them. For example, if a tensor declares it has a constant
region from $i = 2:5$, but some other part of the computation modifies the
tensor at $i = 3$, this would result in incorrect behavior. It is much easier to
write correct Looplet code if we can assume that the tensor is immutable while
we are reading from it. Thus, we introduce the notion that a tensor can be in
read-only mode or update-only mode.  In read-only mode, the tensor may only
appear in the right-hand side of assignments. In update-only mode, the tensor
may only appear in the left-hand side of an assignment, either being overwritten
or incremented by some operator.  We can switch between these modes using freeze
and thaw functions. The declare function is used to allocate a tensor,
initialize it to some specified size and value, and leave it in update-only
mode.

The unfurl function is used to manage iteration over a subfiber. At the point
when it comes time to iterate over a tensor, be in on the left or right hand
side of an assignment, we call the unfurl function to precompute whatever state
and datastructures are necessary to return a looplet nest that would iterate
over that level of a tensor. The unfurl function is called directly before
iterating over the corresponding loop, so it has access to any state variables introduced
by freezing or thawing the tensor.

Our view of a level as a fiber allocator implies an allocation function
$assemble(tns, pos_{start}:pos_{stop})$, which allocates fibers at positions
$pos_{start}:pos_{stop}$ in the level. We don't specify a de-allocation
function, instead relying on initialization to reset the fiber if it needs to be
reused. While all of the previous functions are used to manage the lifecycle and
iteration over a general tensor, the assemble function is quite specific to the
level abstraction, and the notion of positions within sublevels. Note: it was an
intentional choice to hold the parent level responsible for managing the
data of the sublevels, which positions they allocate, etc. This allows the parent
level to reuse allocation logic from internal index datastructures. For example,
a sparse level might use a list of indices to store which nonzeros are present,
and when it comes time to resize that list, it could also call assemble to resize the
sublevel, reducing the number of branches in the code. The assemble function
lends itself particularly to a "vector doubling" allocation approach, which we
have found to be effective and flexible when managing the allocation
of sparse left hand sides. This benefit is made clear in our evaluation section,
where we see that prior systems like TACO do not support all possible loop
orderings and format combinations for sparse matrix multiply because they do
not have a flexible enough allocation strategy, instead using a two-phase approach
which requires computing a complicated closed-form kernel to iterate over the
data twice to determine the number of required output nonzeros.

\begin{figure}
    \raggedright
\paragraph{$declare(lvl, init, dims...)$} Declares the level to hold subtensors
of size $dims$ and an initial value of $init$. Requires the level to be in
read-only mode.
\paragraph{$freeze(lvl)$} Finalizes the updates in the level, and readies the
level for reading. Requires the level to be in update-only mode.
\paragraph{$thaw(lvl)$} Prepares the level to accept updates. Requires the level
to be in read-only mode.
\paragraph{$unfurl(fiber(lvl, pos), ext, mode)$} Unfurls the fiber at position
$pos$ in the level $lvl$ over the extent $ext$. When $mode = \finchread$,
returns a looplet nest over the values in the read-only fiber.  When $mode =
\finchupdate$, returns a looplet nest over mutable subfibers in the update-only
fiber. Often, skipping over mutable locations allows the level to know which
locations must be stored. Additionally, a dirty bit may be used to communicate
whether the mutable subfiber has been written to, which allows the parent fiber
to know whether the subfiber must be stored explicitly.
\paragraph{$assemble(lvl, pos_{start}, pos_{stop})$} Allocates subfibers in the
level from positions $pos_{start}$ to $pos_{stop}$. Usually, this function is
only ever called on unassembled positions, but some levels (such as dense levels
or bytemaps) may support reassembly.
\caption{The five functions that define a level.}
\end{figure}

\subsection{Level Formats}
Having described the 8 basic level structures and the functions that define a
level format, we can now detail the key level structures that we support, some
of the challenges in their implementation, and how we overcome them.

\begin{enumerate}
\item SparseList
\item SparseDict
\item ...
\end{enumerate}
